Hot and fresh: Pervasive climate stressors of seagrass in a large Gulf coast estuary

### 2023-12-08 manu-draft.qmd
text edits to intro, adding cites

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ in areal cover since the 1980s [@waycott2009]. These losses are comparable, if not more, than other critical coastal environments, which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities, whereas natural disturbances such as storm events and disease have also been implicated. Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create suitable conditions supportive for growth. A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients. Cooperation among management, regulatory, and private sectors were critical aspects of both these examples for successful implementation of nutrient controls. In other cases, the reversal of seagrass losses through nutrient reductions may not be possible because of system hysteresis where the path to recovery is not the same as the path to decline. Costly alternative management actions in addition to nutrient reductions are needed in these situations. Management actions are much more effective in estuarine sytems that are stressed but not past critical thresholds that define an alternative state.

@@

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years. Through successful reduction of nitrogen loads by 2/3 from , seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth. Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers. Recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay [@nickerson2023], whereas changing salinity has not been well-described, nor have potential links to seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay Estuary over the last fifty years. Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring data describing cover of individual species, and synoptic seagrass data collected with the water quality data. Specific focus was on the period from prior to and after 2016, when seagrasses recovered baywide and then declined by 28% despite water quality conditions remaining relatively stable, suggesting that additional stressors beyond light requirements may explain the recent decline. The long-term recovery of Tampa Bay as primarily shown through the recovery of seagrasses to 2016 is a nationally recognized success story. This analysis provides a more recent assessment of current conditions by examining climate-related stressors and seagrass loss. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.

>>>>

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services[@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable, if not more, than other critical coastal environments [@duarte2008a], which represents an existential threat to the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @orth2006; @burkholder2007], whereas natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed environments requires management actions that create conditions supportive for growth. A long-standing approach adopted by numerous management entities has been the control of nutrient inputs in systems where excessive phytoplankton growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @greening2014] . There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014]and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in areal coverage through sustained and long-term reductions in nutrients. Cooperation among management, regulatory, and private sectors were critical aspects of both these examples for successful implementation of nutrient controls [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Costly alternative management actions, in addition to nutrient reductions, are needed in these situations. Management actions are much more effective in estuaries that are stressed but not past critical thresholds that define an alternative state.

@@

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years. Through successful reduction of nitrogen loads by 2/3 pre-regulation, seagrasses have recovered from a low in the 1980s to an all-time high in 2016 of 41,655 acres (16,857 ha). Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality with adequate light penetration that has been historically shown to support growth. Factors that have influenced this recent decline are unknown and the effects of climate change on physical conditions have been implicated as potential drivers. Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay [@nickerson2023]. Changing salinity has not been well-described, nor have potential links to seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends and to develop models that can explain relationships of seagrass change with water quality conditions. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

This paper describes a comprehensive assessment of long-term trends in water temperature and salinity in the Tampa Bay Estuary over the last fifty years. Three datasets of varying sample design and temporal coverage were used to assess the primary hypothesis that conditions in Tampa Bay are trending towards hotter and fresher conditions that are likely pushing seagrasses beyond their optimal tolerance ranges. This analysis was supported by 30-year seagrass datasets describing approximate biennial coverage of all seagrasses, annual transect monitoring data describing cover of individual species, and synoptic seagrass data collected with the water quality surveys. Specific focus was on the period from prior to and after 2016, when seagrasses recovered baywide and then declined despite water quality conditions remaining relatively stable. Our primary goal was to relate recent seagrass changes to climate stressors to provide a motivating example for the management community in the Tampa Bay region that a shifting ecological baseline presents new challenges beyond the known success story of long-term bay recovery. The results are also provided as a cautionary example of the subtle but increasing role of climate-related stressors on the resiliency of estuarine systems, prompting a reassessment of existing management paradigms to accommodate current and anticipated future conditions.
